---
title: "Vaginal micro-mycobiome analysis 2025"
author: "Alexandra Roth Schulze"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_notebook:
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_float: yes
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r echo=FALSE}
options(width = 90)
```

```{r ,include=FALSE}
library(knitr)
opts_chunk$set(comment = NA, warning=FALSE, message=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
library("ggplot2")
theme_set(theme_bw())
pal = "Set1"
scale_colour_discrete <- function(palname = pal, ...) {
    scale_colour_brewer(palette = palname, ...)
}
scale_fill_discrete <- function(palname = pal, ...) {
    scale_fill_brewer(palette = palname, ...)
}
```


```{r PackageLoad, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
library("phyloseq")
library("vegan")
```

#Pre-processing

The taxonomic classification was further improved by BLASTing each sequence and assigning a species name based on this in order to agglomerate features into OTUs based on this species assignation. The file containing the information can be found in the folder of this analysis. It will tell to which species each feature had a equally good (same identity %) hit to and to which one corresponds.

Methodology description cited ( doi: 10.7554/eLife.57051 --> https://elifesciences.org/articles/57051) the authors describe as part of their pipeline:

For the three most dominant ASVs in each vaginal CST, we performed BLASTN (v 2.10.1+) against the NCBI 16S ribosomal RNA database (Zhang et al., 2000), excluding models (XM/XP) and uncultured/environmental samples (Source data 2). Alignments were filtered to minimum 99% identity, and the ASVs were assigned to unique species, when possible. In these cases where multiple species aligned equally well, we used published studies to select the most likely species. As an example, L. crispatus and L. acidophilus could not be distinguished based on the sequenced region, and as published studies on the vaginal microbiota concur that L. crispatus, in contrast to L. acidophilus, constitute an important part of the vaginal microbiota, we refer to L. crispatus/L. acidophilus solely as L. crispatus. Based on this reasoning, we will also refer to L. gasseri/johnsonii solely as L. gasseri (Antonio et al., 1999; Kiss et al., 2007; Vásquez et al., 2002).

#Microbiome

```{r Imp.format, echo=FALSE}
###Importing and formating microbiome
setwd("~/Documents/ENDIA Project/Data and analyses/ENDIA Projects analysis/2025 Sequencing processing with qiime2 2024.10/Vaginal project/") # HERE YOU SET YOUR DIRECTORY WHERE DATA IS CONTAINED
Feat <- load(file="./Bacterial_Vaginal_Features_Phyloseq_Obj_Filtered1.RData")
#Bacterial_VCMCFHS_AFilt1
  
#CHECK HERE
#New classification (from Kingdom to family), species and genus classifications will be reassigned as per BLAST n below
tax_table(Bacterial_VCMCFHS_AFilt1)["7c942fbcc204080381806f4cc91f41f2",] <- tax_table(Bacterial_VCMCFHS_AFilt1)["9002ee8e03d4dfdf26665209e354ce2f",]
tax_table(Bacterial_VCMCFHS_AFilt1)["9df9743e450924380cd7a4d3cd9c8a59",] <- tax_table(Bacterial_VCMCFHS_AFilt1)["9002ee8e03d4dfdf26665209e354ce2f",]
tax_table(Bacterial_VCMCFHS_AFilt1)["7d35d4541ed0fb016dad07caad37f556",] <- tax_table(Bacterial_VCMCFHS_AFilt1)["9002ee8e03d4dfdf26665209e354ce2f",]
tax_table(Bacterial_VCMCFHS_AFilt1)["297edc18fabd1d1b3234f8ef062a36c2",] <- tax_table(Bacterial_VCMCFHS_AFilt1)["e716d8c424a91dc74faa15671716cfc1",]
tax_table(Bacterial_VCMCFHS_AFilt1)["30d3ed513d2487ebacec1d3806b5ccae",] <- tax_table(Bacterial_VCMCFHS_AFilt1)["563395e43ea73827809d66727b023e5b",]
tax_table(Bacterial_VCMCFHS_AFilt1)["30d3ed513d2487ebacec1d3806b5ccae", "Family"] <- "Lawsonellaceae"

NewClass <- read.table(file="./BLASTn_classification_vaginal_microbiome.csv", header=TRUE, row.names=1, sep="\t")
NewClass[,"Good_Class"] <- as.character(NewClass[,"Good_Class"])
NewClass[,"Genus"] <- as.character(NewClass[,"Genus"])
colnames(tax_table(Bacterial_VCMCFHS_AFilt1))[1] <- "Kindom"
rownames(NewClass) <- NewClass[,"OTU_names"]
 
Bacterial_VCMCFHS_AFilt12 <- Bacterial_VCMCFHS_AFilt1
new_order <- sort(taxa_names(Bacterial_VCMCFHS_AFilt12)) 
# Reorder the OTU table
Reordered_ASV <- otu_table(Bacterial_VCMCFHS_AFilt12)[,new_order]
# Reorder the taxonomy table
Reordered_Tax <- tax_table(Bacterial_VCMCFHS_AFilt12)[new_order, ]

#Make phyloseq object
Reffasta <- refseq(Bacterial_VCMCFHS_AFilt12)
Meta <- sample_data(Bacterial_VCMCFHS_AFilt12)
Bacterial_VCMCFHS_AFilt13 <- phyloseq(otu_table(Reordered_ASV, taxa_are_rows = F),  tax_table(Reordered_Tax), sample_data(Meta), refseq(Reffasta))

tax_table(Bacterial_VCMCFHS_AFilt13)[,"Species"] <- NewClass[,"Good_Class"]
tax_table(Bacterial_VCMCFHS_AFilt13)[,"Genus"] <- NewClass[,"Genus"]
OTUVag <- tax_glom(Bacterial_VCMCFHS_AFilt13, taxrank = "Species", NArm=FALSE)
OTUVag <-prune_samples(sample_sums(OTUVag)>=1000, OTUVag)
 #Current phyla
#"Bacteroidota"     "Firmicutes"       "Actinobacteriota" "Proteobacteria"  "Fusobacteriota"   "Campilobacterota"

#Genus
OTUVag_G <- tax_glom(OTUVag, taxrank = rank_names(OTUVag_G)[6], NArm=FALSE)
taxa_names(OTUVag_G) <- tax_table(OTUVag_G)[,"Genus"]

#Family
OTUVag_F <- tax_glom(OTUVag, taxrank = "Family", NArm=FALSE)
taxa_names(OTUVag_F) <- tax_table(OTUVag_F)[,"Family"]

#Order
OTUVag_O <- tax_glom(OTUVag, taxrank = "Order", NArm=FALSE)
taxa_names(OTUVag_O) <- tax_table(OTUVag_O)[,"Order"]

#Phylum
OTUVag_P <- tax_glom(OTUVag, taxrank = "Phylum", NArm=FALSE)
taxa_names(OTUVag_P) <- tax_table(OTUVag_P)[,"Phylum"]

VecO <- list("Species" = OTUVag, "Genus" = OTUVag_G, "Family"= OTUVag_F, "Order"= OTUVag_O, "Phylum"= OTUVag_P)

#Rarefy
OTUVag_RA <- rarefy_even_depth(OTUVag, sample.size = 7000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Genus
OTUVag_RAG <- tax_glom(OTUVag_RA, taxrank = "Genus", NArm=FALSE)
taxa_names(OTUVag_RAG) <- tax_table(OTUVag_RAG)[,"Genus"]

#Family
OTUVag_RAF <- tax_glom(OTUVag_RA, taxrank = "Family", NArm=FALSE)
taxa_names(OTUVag_RAF) <- tax_table(OTUVag_RAF)[,"Family"]

#Order
OTUVag_RAO <- tax_glom(OTUVag_RA, taxrank = "Order", NArm=FALSE)
taxa_names(OTUVag_RAO) <- tax_table(OTUVag_RAO)[,"Order"]

#Phylum
OTUVag_RAP <- tax_glom(OTUVag_RA, taxrank = "Phylum", NArm=FALSE)
taxa_names(OTUVag_RAP) <- tax_table(OTUVag_RAP)[,"Phylum"]

#Dummy
TLevel<-c("Species", "Genus", "Family", "Order", "Phylum")

Vec <- list("Species" = OTUVag_RA, "Genus" = OTUVag_RAG, "Family"= OTUVag_RAF, "Order"= OTUVag_RAO, "Phylum"= OTUVag_RAP)
Vec2 <- list("Species" = OTUVag_RA, "Genus" = OTUVag_RAG, "Family"= OTUVag_RAF, "Order"= OTUVag_RAO, "Phylum"= OTUVag_RAP)

for (i in 1:5) {
#Dummy <- Vec[[i]]
Dummy <- VecO[[i]]
#Making sure each factor has the correct class
#This factor corresponds to extraction batch
sample_data(Dummy)$SeqRun <- as.factor(sample_data(Dummy)$SeqRun)
#This factor corresponds to gestational age (day into pregnancy in which the sample was taken)
sample_data(Dummy)$Days <- as.integer(as.character(sample_data(Dummy)$Days))
#This factor corresponds to conception age
sample_data(Dummy)$Age <- as.numeric(as.character(sample_data(Dummy)$Age))
#This factor corresponds to Body Mass Index at conception time
sample_data(Dummy)$BMI <- as.numeric(as.character(sample_data(Dummy)$BMI))
#This factor corresponds to Human leukocyte antigen (HLA) type of the women
sample_data(Dummy)$HLA <- as.factor(sample_data(Dummy)$HLA)
#This factor corresponds to parity (which refers to if the women already had children before the current pregnancy)
sample_data(Dummy)$Parity <- as.factor(sample_data(Dummy)$Parity)
#This factor corresponds to a unique identifier per person (i.e. if the woman contributed samples from different pregnancies, those samples belong to the same motherid)
sample_data(Dummy)$motherid <- as.factor(sample_data(Dummy)$motherid)
#This factor corresponds to T1D status
sample_data(Dummy)$T1Dstatus <- as.factor(sample_data(Dummy)$T1Dstatus)
#This factor corresponds to trimester
sample_data(Dummy)$TriCorr <- as.factor(sample_data(Dummy)$TriCorr)

#Vec2[[i]] <- Dummy
VecO[[i]] <- Dummy
}
#VecO --> Is the data without rarefaction
#Vec2 --> Is the Rarefied data
#Sample without HLA info should be removed when this factor i included in the statistical models in later analyses
Bac <- VecO
Bac_RA <- Vec2
save(Bac, Bac_RA, file="./MicrobiomeObjects.RData")
```

#Mycobiome 

```{r Myco, echo=FALSE, fig.height=12, fig.width=16}
##Importing mycobiome
Feat <- load(file="./Myco_ITS1_R1_CU_Filter1.RData")
#Myco_ITS1_R1_CU_Filter1

#phyloseq-class experiment-level object
#otu_table()   OTU Table:         [ 593 taxa and 153 samples ]
#sample_data() Sample Data:       [ 153 samples by 11 sample variables ]
#tax_table()   Taxonomy Table:    [ 593 taxa by 8 taxonomic ranks ]
#phy_tree()    Phylogenetic Tree: [ 593 tips and 591 internal nodes ]
#refseq()      DNAStringSet:      [ 593 reference sequences ] 

#Note, before there were also 153 samples at this point

##FILTER SEQUENCES CLASSIFIED AS UNASSIGNED
Myco_ITS1_R1_CU_Filter1_F <- subset_taxa(Myco_ITS1_R1_CU_Filter1, Kindom == "k__Fungi")
#This doesn't remove any feature or sample
Myco_ITS1_R1_CU_Filter1_F <- subset_samples(Myco_ITS1_R1_CU_Filter1_F, SampleType == "Vagina")

sample_names(Myco_ITS1_R1_CU_Filter1_F) <- sample_data(Myco_ITS1_R1_CU_Filter1_F)$Merging

NewMet <- read.table(file="./Metadata_8_17_Merged2PET1D.csv", header=TRUE, row.names=1, sep="\t")

Myco_ITS1_R1_CU_Filter1_F <- phyloseq(otu_table(Myco_ITS1_R1_CU_Filter1_F), tax_table(Myco_ITS1_R1_CU_Filter1_F), phy_tree(Myco_ITS1_R1_CU_Filter1_F), refseq(Myco_ITS1_R1_CU_Filter1_F), sample_data(NewMet))

#phyloseq-class experiment-level object
#otu_table()   OTU Table:         [ 593 taxa and 151 samples ]
#sample_data() Sample Data:       [ 151 samples by 11 sample variables ]
#tax_table()   Taxonomy Table:    [ 593 taxa by 8 taxonomic ranks ]
#phy_tree()    Phylogenetic Tree: [ 593 tips and 591 internal nodes ]
#refseq()      DNAStringSet:      [ 593 reference sequences ]

##Make factor
sample_data(Myco_ITS1_R1_CU_Filter1_F)$T1Dstatus <- as.factor(sample_data(Myco_ITS1_R1_CU_Filter1_F)$T1Dstatus)
sample_data(Myco_ITS1_R1_CU_Filter1_F)$HLA <- as.factor(sample_data(Myco_ITS1_R1_CU_Filter1_F)$HLA)

#We will be working with original ASVs/features and with data agglomerated to "Species Hypotheses" level
Myco_ITS1_R1_CU_Filter1_F_GSH <- tax_glom(Myco_ITS1_R1_CU_Filter1_F, taxrank = "Sp.Hypotheses.", NArm=FALSE) # Definition: SHs are clusters of fungal DNA sequences that are considered to represent a species or species-like unit. These clusters are determined using sequence similarity thresholds applied to marker genes, such as the ITS region, which is the primary fungal DNA barcode.

# Convert names that are not useful to NA
#tax_table(DummyTOTUF) =gsub("uncultured organism", as.character(NA), tax_table(DummyTOTUF))

#rarefy to 2000 sequences per sample --> Only one of the current samples has less than 2000 sequences, the same in both datasets
Myco_ITS1_R1_CU_Filter1_F_RA <- rarefy_even_depth(Myco_ITS1_R1_CU_Filter1_F, sample.size = 2000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
Myco_ITS1_R1_CU_Filter1_F_RA5000 <- rarefy_even_depth(Myco_ITS1_R1_CU_Filter1_F, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

Myco_ITS1_R1_CU_Filter1_F_GSH_RA <- rarefy_even_depth(Myco_ITS1_R1_CU_Filter1_F_GSH, sample.size = 2000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
Myco_ITS1_R1_CU_Filter1_F_GSH_RA5000 <- rarefy_even_depth(Myco_ITS1_R1_CU_Filter1_F_GSH, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Genus
My_G <- tax_glom(Myco_ITS1_R1_CU_Filter1_F, taxrank = "Genus", NArm=TRUE)
taxa_names(My_G) <- tax_table(My_G)[,"Genus"]
My_G_RA5000 <- rarefy_even_depth(My_G, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Family
My_F <- tax_glom(Myco_ITS1_R1_CU_Filter1_F, taxrank = "Family", NArm=TRUE)
taxa_names(My_F) <- tax_table(My_F)[,"Family"]
My_F_RA5000 <- rarefy_even_depth(My_F, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Order
My_O <- tax_glom(Myco_ITS1_R1_CU_Filter1_F, taxrank = "Order", NArm=TRUE)
taxa_names(My_O) <- tax_table(My_O)[,"Order"]
My_O_RA5000 <- rarefy_even_depth(My_O, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Phylum
My_P <- tax_glom(Myco_ITS1_R1_CU_Filter1_F, taxrank = "Phylum", NArm=TRUE)
taxa_names(My_P) <- tax_table(My_P)[,"Phylum"]
My_P_RA5000 <- rarefy_even_depth(My_P, sample.size = 5000, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

Myco <- list("ASV" = Myco_ITS1_R1_CU_Filter1_F, "Species" = Myco_ITS1_R1_CU_Filter1_F_GSH, "Genus" = My_G, "Family"= My_F, "Order"= My_O, "Phylum"= My_P)
Myco_RA5000 <- list("ASV" = Myco_ITS1_R1_CU_Filter1_F_RA5000, "Species" = Myco_ITS1_R1_CU_Filter1_F_GSH_RA5000, "Genus" = My_G_RA5000, "Family"= My_F_RA5000, "Order"= My_O_RA5000, "Phylum"= My_P_RA5000)

#Genus
My_G_RA <- tax_glom(Myco_ITS1_R1_CU_Filter1_F_RA, taxrank = "Genus", NArm=TRUE)
taxa_names(My_G_RA) <- tax_table(My_G_RA)[,"Genus"]

#Family
My_F_RA <- tax_glom(Myco_ITS1_R1_CU_Filter1_F_RA, taxrank = "Family", NArm=TRUE)
taxa_names(My_F_RA) <- tax_table(My_F_RA)[,"Family"]

#Order
My_O_RA <- tax_glom(Myco_ITS1_R1_CU_Filter1_F_RA, taxrank = "Order", NArm=TRUE)
taxa_names(My_O_RA) <- tax_table(My_O_RA)[,"Order"]

#Phylum
My_P_RA <- tax_glom(Myco_ITS1_R1_CU_Filter1_F_RA, taxrank = "Phylum", NArm=TRUE)
taxa_names(My_P_RA) <- tax_table(My_P_RA)[,"Phylum"]

Myco_RA <- list("ASV" = Myco_ITS1_R1_CU_Filter1_F_RA, "Species" = Myco_ITS1_R1_CU_Filter1_F_GSH_RA, "Genus" = My_G_RA, "Family"= My_F_RA, "Order"= My_O_RA, "Phylum"= My_P_RA)

save(Myco, Myco_RA, Myco_RA5000, file="./MycobiomeObjects.RData")
```
